---
title: "Causal inference / SEED"
author: "Amy Cochran"
format:
  html:
    css: taskiia.css
    theme: cosmo
    toc: true
    toc-depth: 3
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This notebook estimates the average treatment effect (ATE) of several exposuresŌĆöeach considered individuallyŌĆöon autism case status as defined in the SEED dataset. All exposures and covariates reflect information available at or around birth. We use Bayesian Additive Regression Trees (BART) to flexibly model the outcome, allowing for nonlinearities and interactions without overfitting.

For each exposure, we adjust for a consistent set of potential confounders, including both a predefined covariate list and the other exposures. Because the exposures share a common set of potential confounders (including one another), we use the same adjustment strategy across all models. This allows each exposureŌĆÖs effect to be estimated while accounting for shared sources of confounding, ensuring fair comparisons and consistent interpretation.

# Initialization

First, we source an external R script that loads all necessary libraries and helper functions. Next, we load a configuration file written in YAML format, which specifies important analysis settings such as file paths, variable names, and modeling options. Finally, we read in the raw dataset specified in the configuration file, which will be cleaned and processed in the next steps.

```{r}
# Load necessities
source("imports.R")

# Load configuration
config <- yaml::read_yaml("config.yaml")

# Load data
raw_df <- readr::read_csv(config$data_file)
```

The table below summarizes the exposures, covariates, and outcome used in the analysis, based on the configuration file.

```{r}
# 
# Extract exposure names and types
exposures_df <- do.call(rbind, lapply(config$exposures, as.data.frame))
exposures_df$category <- "Exposure"

# Extract covariates
covariates_df <- do.call(rbind, lapply(config$covariates, as.data.frame))
covariates_df$category <- "Covariate"

# Extract outcome
outcome_df <- as.data.frame(config$outcome)
outcome_df$category <- "Outcome"

# Combine all
var_summary <- rbind(exposures_df, covariates_df, outcome_df[, c("name", "type","category")])
names(var_summary) <- c("Name", "Type", "Category")

# Reorder columns
var_summary <- var_summary[, c("Category", "Name", "Type")]

# Render as markdown table
knitr::kable(var_summary, caption = "Variables defined in the analysis configuration.")

```

# Data preparation

## Coerce data type

We begin by preparing the analysis dataset. First, we extract only the variables relevant for our analysisŌĆöspecifically, the exposures, covariates, and outcome specified in the configuration. We then coerce the data into the correct type, either continuous or categorical.

```{r}

# Initialize dataframe for coercion
coerced_df <- raw_df

# Keep only exposures, covariates, and outcome
vars_to_keep <- c(
  unlist(lapply(config$exposures, function(x) x$name)),
  unlist(lapply(config$covariates, function(x) x$name)),
  config$outcome$name
)
coerced_df <- coerced_df[, intersect(vars_to_keep, names(coerced_df))]

# Combine exposure, covariate, and outcome types into a named vector
type_map <- c(
  setNames(vapply(config$exposures, `[[`, character(1), "type"),
           vapply(config$exposures, `[[`, character(1), "name")),
  setNames(vapply(config$covariates, `[[`, character(1), "type"),
           vapply(config$covariates, `[[`, character(1), "name")),
  setNames(config$outcome$type, config$outcome$name)
)

# Coerce variable types in analytic_df to match type_map
for (v in names(type_map)) {
  if (!v %in% names(coerced_df)) next
  coerced_df[[v]] <- switch(
    type_map[[v]],
    continuous  = as.numeric(coerced_df[[v]]),
    categorical = as.factor(coerced_df[[v]]),
    stop("Unsupported type: ", type_map[[v]])
  )
}

# Coerce outcome data into 0's and 1's
if (config$outcome$type == "continuous") {
  # Leave as-is
  coerced_df[[config$outcome$name]] <- coerced_df[[config$outcome$name]]
} else {
  # Relevel and convert to 0/1
  f <- relevel(factor(as.character(coerced_df[[config$outcome$name]])),
               ref = as.character(config$outcome$reference))
  coerced_df[[config$outcome$name]] <- as.integer(f) - 1
}

# Identify continuous exposures
continuous_exposures <- vapply(
  config$exposures,
  function(x) if (x$type == "continuous") x$name else NA_character_,
  character(1)
)
continuous_exposures <- na.omit(continuous_exposures)

# Compute mean and SD on original (pre-coercion) data
exposure_stats <- lapply(continuous_exposures, function(var) {
  vals <- coerced_df[[var]]
  data.frame(
    exposure = var,
    mean     = mean(vals, na.rm = TRUE),
    sd       = sd(vals, na.rm = TRUE)
  )
})

# Combine into single data.frame
exposure_stats_df <- do.call(rbind, exposure_stats)
summary(exposure_stats_df)

# Summarize data
summary(coerced_df)

```
## Add missingness indicators

For a small set of variables of interest, we generate missingness indicators to track whether a value was missing, and we replace missing values with zero for those variables to preserve information about potential patterns of missingness. 

```{r}

# Initialize dataframe for indicators
indicated_df <- coerced_df

# ---- Create missingness indicators and update config ----
for (var in config$missingness_indicators) {
  miss_name <- paste0("miss_", var)
  
  # Create indicator column
  indicated_df[[miss_name]] <- as.integer(is.na(indicated_df[[var]]))
  
  # Replace missing values in original variable with 0
  indicated_df[[var]][is.na(indicated_df[[var]])] <- 0

  # Add missingness indicator to config under correct list
  entry <- list(name = miss_name, type = "binary")
  
  if (var %in% sapply(config$exposures, `[[`, "name")) {
    config$exposures <- append(config$exposures, list(entry))
  } else if (var %in% sapply(config$covariates, `[[`, "name")) {
    config$covariates <- append(config$covariates, list(entry))
  } else {
    warning(paste("Variable", var, "not found in exposures or covariates."))
  }
}

# Summarize data 
summary(indicated_df)

```

## Impute remaining missingness

We perform single imputation on the remaining missing values to create a complete dataset for analysis.


```{r}

# Initialize dataframe for imputation
imputed_df <- indicated_df

# Temporarily remove outcome
outcome_name <- config$outcome$name
outcome <- imputed_df[[outcome_name]]
imputed_df[[outcome_name]] <- NULL

# Convert character columns to factors
char_vars <- sapply(imputed_df, is.character)
imputed_df[char_vars] <- lapply(imputed_df[char_vars], factor)

# Impute missing data
imputed_df <- complete(mice(imputed_df, m = 1, print = TRUE))

# Add outcome back
imputed_df[[outcome_name]] <- outcome

# Summarize data
summary(imputed_df)

```

## Create dummy variables

Finally, we convert categorical variables to dummy variables so they can be used in models that require numeric inputs. In each case, we drop the reference category to avoid collinearity, ensuring the most common category serves as the baseline. 


```{r}

# Initialize dataframe for final step
analytic_df <- imputed_df

# Create dummy variables and update config
new_covariates <- list()
new_exposures  <- list()

for (var in c(config$covariates, config$exposures)) {
  if (var$type == "categorical") {
    var_name <- var$name
    
    # Relevel with most frequent category as reference
    most_common <- names(sort(table(analytic_df[[var_name]]), decreasing = TRUE))[1]
    message("Most common level for ", var_name, ": ", most_common)
    analytic_df[[var_name]] <- relevel(factor(analytic_df[[var_name]]), ref = most_common)

    # Create dummy variables, dropping reference
    dummies <- model.matrix(~ . - 1, data = analytic_df[var_name])
    ref_colname <- paste0(var_name, most_common)
    dummies <- dummies[, colnames(dummies) != ref_colname, drop = FALSE]

    # Add to analytic_df
    analytic_df <- bind_cols(analytic_df, as.data.frame(dummies))
    analytic_df[[var_name]] <- NULL

    # Decide if the variable came from exposures or covariates
    is_exposure <- var_name %in% sapply(config$exposures, function(x) x$name)

    # Build entries
    dummy_names <- colnames(dummies)
    dummy_entries <- lapply(dummy_names, function(name) list(name = name, type = "binary"))

    if (is_exposure) {
      new_exposures <- append(new_exposures, dummy_entries)
    } else {
      new_covariates <- append(new_covariates, dummy_entries)
    }
  }
}

# Drop variables with missing outcomes
# Count missing outcome values
n_missing <- sum(is.na(analytic_df[[config$outcome$name]]))

# Drop rows with missing outcome
analytic_df <- analytic_df[!is.na(analytic_df[[config$outcome$name]]), ]

# Print message in Quarto
cat(sprintf("Dropped %d row%s with missing outcome (%s).\n",
            n_missing,
            ifelse(n_missing == 1, "", "s"),
            config$outcome$name))

# Initialize finalized configuration file
fin_config <- config

# Remove original categorical variables from config
fin_config$covariates <- Filter(function(x) !(x$type == "categorical"), fin_config$covariates)
fin_config$exposures  <- Filter(function(x) !(x$type == "categorical"), fin_config$exposures)

# Add new dummy variables to config
fin_config$covariates <- append(fin_config$covariates, new_covariates)
fin_config$exposures  <- append(fin_config$exposures, new_exposures)

summary(analytic_df)
```
Let's check the updated list of covariates, exposures, and outcomes, now that we did a little pre-processing:

The table below summarizes the exposures, covariates, and outcome used in the analysis, based on the configuration file.

```{r}
# 
# Extract exposure names and types
exposures_df <- do.call(rbind, lapply(fin_config$exposures, as.data.frame))
exposures_df$category <- "Exposure"

# Extract covariates
covariates_df <- do.call(rbind, lapply(fin_config$covariates, as.data.frame))
covariates_df$category <- "Covariate"

# Extract outcome
outcome_df <- as.data.frame(fin_config$outcome)
outcome_df$category <- "Outcome"

# Combine all
var_summary <- rbind(exposures_df, covariates_df, outcome_df[, c("name", "type","category")])
names(var_summary) <- c("Name", "Type", "Category")

# Reorder columns
var_summary <- var_summary[, c("Category", "Name", "Type")]

# Render as markdown table
knitr::kable(var_summary, caption = "Variables defined in the analysis configuration.")

```

# Estimating ATE for Binary Exposures

To estimate the average treatment effect (ATE) of each binary exposure, we use Bayesian Additive Regression Trees (BART) to flexibly model the outcome as a function of the exposure and other covariates (including other exposures). BART accommodates nonlinearities and complex interactions without requiring them to be specified in advance, making it well suited for causal effect estimation.

As part of the modeling process, we also fit a treatment model to obtain an estimated propensity score, which we include as an additional covariate in the response model. While this score is not essential, including it can reduce bias in ATE estimates without constraining the model's flexibility. This approach is supported by prior work showing that BARTŌĆÖs performance often improves when provided with this additional information (see Hill, 2011 and Hahn et al., 2020).

The ATE is then computed as the posterior average difference in predicted outcomes under treatment versus control, with 95% credible intervals constructed from the posterior samples.

```{r}
# Initialize results storage
ate_results <- data.frame(
  exposure = character(),
  ate = numeric(),
  ci_lower = numeric(),
  ci_upper = numeric(),
  stringsAsFactors = FALSE
)

# Loop over binary exposures
for (exposure_info in fin_config$exposures) {
  if (!identical(exposure_info$type, "binary")) next
  
  message("Processing exposure: ", exposure_info$name)

  
  exposure_var <- exposure_info$name
  outcome_var  <- fin_config$outcome$name
  
  covariates <- setdiff(names(analytic_df), c(exposure_var, outcome_var))
  
  # Fit: BART for treatment + outcome
  fit <- bartc(
    response     = analytic_df[outcome_var],
    treatment    = analytic_df[exposure_var],
    confounders  = analytic_df[, covariates],
    estimand     = "ate",
    method.trt   = "bart",
    method.rsp   = "bart",
    commonSup.rule = "none"
  )
  
  # Summarize fit
  summary_fit <- summary(fit)
  
  # Save results
  ate_results <- rbind(
    ate_results,
    data.frame(
      exposure  = exposure_var,
      ate       = summary_fit$estimates$estimate,
      ci_lower  = summary_fit$estimates$ci.lower,
      ci_upper  = summary_fit$estimates$ci.upper,
      stringsAsFactors = FALSE
    )
  )
  
  

}
```

# Estimating ATE for continuous exposures

We estimate the ATE for each continuous exposure using a two-step BART approach. Since bartc() is limited to binary or categorical exposures, we first model the exposure using BART to obtain its posterior mean, treating it as a function of all covariates. Assuming normally distributed, homoskedastic residuals, we construct a generalized propensity score (GPS), up to a transformation, as the squared deviation between observed and predicted exposure values. This GPS is added as a covariate in a second BART model for the outcome.

To estimate the ATE, we compare predicted outcomes under two hypothetical exposure valuesŌĆöone standard deviation above and below the meanŌĆöwhile holding all other variables fixed. This contrast yields the estimated effect of increasing the exposure from a low to high level. This approach builds on the framework proposed by Hill (2011) and extended in Hahn et al. (2020), where modeling flexibility and covariate adjustment are prioritized over direct modeling of the treatment effect surface.

```{r}

# Loop over binary exposures
for (exposure_info in fin_config$exposures) {
  if (!identical(exposure_info$type, "continuous")) next
  
  message("Processing exposure: ", exposure_info$name)

  # Exposure and outcome variable name
  exposure_var <- exposure_info$name
  outcome_var  <- fin_config$outcome$name
  
  # List of control variables (covariates + remaining exposures)
  covariates <- setdiff(names(analytic_df), c(exposure_var, outcome_var))
  
  # Build a model for treatment
  trt_fit <- dbarts::bart(
  x.train = analytic_df[, covariates],
  y.train = analytic_df[[exposure_var]],
  keeptrees = TRUE,
  verbose = FALSE,
  nchain = 10
  )

  # Get posterior mean of treatment model
  samples <- dbarts::extract(trt_fit, combineChains = TRUE)
  mu_x    <- colMeans(samples)
  
  # Generalized propensity score 
  # (transformed under normal + homoskedasticity assumptions)
  propensity_scores <- (analytic_df[[exposure_var]] - mu_x)^2
  
  # Build a new dataset with propensity scores added
  new_df     <- analytic_df
  new_df$propensity_scores <- propensity_scores
  
  # Build testing dataset (mu +/- 1SD)
  mu <- exposure_stats_df$mean[exposure_stats_df$exposure == exposure_var]
  sd <- exposure_stats_df$sd[exposure_stats_df$exposure == exposure_var]
  x_test_high <- new_df
  x_test_high[[exposure_var]] <- mu + sd
  x_test_low <- new_df
  x_test_low[[exposure_var]] <- mu - sd
  
  # Fit response model
  rsp_fit <- dbarts::bart(
  x.train = new_df[, c(covariates, exposure_var, "propensity_scores")],
  y.train = new_df[[outcome_var]],
  x.test  = rbind(
    x_test_high[, c(covariates, exposure_var, "propensity_scores")],
    x_test_low[,  c(covariates, exposure_var, "propensity_scores")]
  ),
  keeptrees = TRUE,
  verbose = FALSE,
  nchain = 10L
 )
  
  # Summarize fit
  n_test <- ncol(rsp_fit$yhat.test) / 2  # should be 5751

  # Take the difference between +1SD and -1SD predictions
  diffs <- rsp_fit$yhat.test[, 1:n_test] - 
           rsp_fit$yhat.test[, (n_test + 1):(2 * n_test)]

  # Compute the mean difference across all test observations
  ate_samples <- colMeans(diffs)

  # Compute posterior mean and 95% CI using normal approximation
  ate_est <- mean(ate_samples)
  ate_se  <- sd(ate_samples)
  ci_lower <- ate_est - 1.96 * ate_se
  ci_upper <- ate_est + 1.96 * ate_se
  
  # Save results
  ate_results <- rbind(
    ate_results,
    data.frame(
      exposure  = exposure_var,
      ate       = ate_est,
      ci_lower  = ci_lower,
      ci_upper  = ci_upper,
      stringsAsFactors = FALSE
    )
  )
}
```

# Save results

To prepare publication-ready results, we summarize the estimated ATEs and their 95% confidence intervals by creating a formatted column in the style estimate (lower CI, upper CI). We extract the exposure name and formatted estimate into a clean table, relabel columns for readability, and save the results as a CSV. We also render the table using knitr::kable() for direct inclusion in a Quarto or R Markdown report.

```{r}
# ---- Build contrast string from exposure_stats_df ----

get_contrast <- function(exposure) {
  row <- exposure_stats_df[exposure_stats_df$exposure == exposure, ]
  
  if (nrow(row) == 0) return("1 vs. 0")
  
  row <- row[1, ]  # just in case
  
  if (is.na(row$mean) || is.na(row$sd)) return("contrast not found")
  
  hi <- round(row$mean + row$sd, 2)
  lo <- round(row$mean - row$sd, 2)
  sprintf("%.2f vs %.2f", hi, lo)
}

# Apply to each row of ate_results
ate_results$contrast <- vapply(ate_results$exposure, get_contrast, FUN.VALUE = character(1))

# Format ATE and CI
ate_results$estimate_ci <- sprintf(
  "%.2f (%.2f, %.2f)",
  round(ate_results$ate, 2),
  round(ate_results$ci_lower, 2),
  round(ate_results$ci_upper, 2)
)

# Final table
table_for_pub <- ate_results[, c("exposure", "contrast", "estimate_ci")]
names(table_for_pub) <- c("Exposure", "Contrast", "ATE (95% CI)")

# Save
write.csv(table_for_pub, "ate_table.csv", row.names = FALSE)

# Markdown or LaTeX table
knitr::kable(
  table_for_pub,
  caption = "Estimated average treatment effects (ATE) and 95% confidence intervals. Contrasts reflect 1 SD above vs. below the mean for continuous exposures, and 1 vs 0 for binary exposures."
)
```
